Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2020-11-24 manuscript_Huang.tex
new figures included.

As a first step, we consider four representative regimes that can be observed at particular values of the control parameter $r$, corresponding to either periodic dynamics or chaotic dynamics close to band merging, in some laminar state, and at the outer crisis. In all examples, we will consistently use the embedding parameters $D = \tau = 6$. In general, we emphasize that a higher value of $D$ leads to a larger variety of ordinal patterns and, hence, a more robust statistics. On the other hand, for evaluating the statistical properties of either pattern frequencies or transition frequencies, this comes on the cost of requiring increasingly longer time series. In the following, we will use realizations of the logistic map comprising \textcolor{red}{$L=xxx$} iterations. \textcolor{red}{Yong: please add the used time series length! Is it the same as in the next subsection...?}

@@

Due to our choice of $\tau$ and $D$, all SCMs take values of zero for the period-3 series because the temporal distance between each pair of components of the embedding vector covers a multiple of two full periods (cf.~Fig.~\ref{fig:embed}). Other choices of $D$ and $\tau$ would commonly yields non-zero values for the period-3 regime. For the three other cases of band merging, laminar state and outer crisis, the values of all SCMs significantly differ from zero. The pattern frequency based SCM $\mathcal{C}_O$ shows smaller values than the other two measures for all regimes, which implies a positioning in the lower right parts of the complexity--entropy planes in Fig.~\ref{fig:fBmNoise}(a). However, for the transition matrix based SCMs, both the estimated entropies $\mathcal{H}_W$, $\mathcal{H}_E$ and associated complexity measures $\mathcal{C}_W$, $\mathcal{C}_E$ show results that are more consistent with the expectations when considering the respective level of chaoticity of the system in those three regimes (i.e., a rising Lyapunov exponent with increasing $r$ in the respective chaotic regimes). In particular, the estimated SCMs take their largest values for $r = 4$ in the outer crisis regime, and the SCMs are the smallest when $r = 3.679$ in the band merging case, while the laminar regime displays intermediate SCM values.

@@

In order to put our previous results into a broader context, we next show the complexity--entropy planes obtained when varying the control parameter $r$ in the logistic map. It is notable that the logistic map has already been widely discussed in the framework of SCMs \cite{RossoPRE2007,MartinPLA2003}, but only for pattern frequency based complexity measures. For each value of $r$, we construct an OPTN from a time series of length $L = 31000$. In order to exclude transient behavior, we remove the first $1000$ values from the data series. We have repeated the following calculations for embedding parameters $D = \tau \in [2, 7]$.

When only one unique ordinal pattern is identified in a given time series, all SCMs take a value of $0$, which is reasonable since only self-loops are observed. This however happens only if $D$ is an integer multiple of the period for the periodic windows of the logistic map (e.g., Fig.~\ref{fig:embed}(g)). Another practical concern is the data requirements for computing entropies and complexity measures. For a reliable estimation of SCMs, a pragmatic condition on the length $L$ of the time series $L$ as being sufficiently larger than $D!$ \cite{rossoPRL2007,kowalskiPhyD2007} has been satisfied in our setting for $D \leq 7$. In the particular example of logistic map, one may easily generate realizations of longer lengths, which however do not improve the obtained estimates significantly (as we have tested for $L\leq 5 \times 10^6$, not shown).

@@

Figure~\ref{fig:fBmNoise} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar when $D$ and $\tau$ are varied in the interval $[3, 7]$. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. By contrast, $\mathcal{C}_{min}$ shows less dependence on $D$.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a,d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b,e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c,f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The dashed lines correspond to the maximum $\mathcal{C}_{max}$ and minimum complexity $\mathcal{C}_{min}$ values for a fixed value of the entropy. Panels (a-c) show the results for embedding dimensions $D = 4,5$ and (d-f) for $D = 6$. In panels (a,d), the bottom-right rectangles of dotted red lines highlight the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. \textcolor{red}{I don't quite get where this information comes from. Is it taken from Rosso's works or own results of this study? I do not see stochastic processes like fBm/fGn being discussed here, which is probably also not necessary. But we should clarify this.}}  \label{fig:fBmNoise}}
\end{figure*}

First of all, we recover the known result that pattern frequency based SCM values lie around the maximum curve $\mathcal{C}_{max}$, while the displacements of the estimated values from their theoretical upper bounds depend on the embedding dimension $D$ as shown in Fig.~\ref{fig:fBmNoise}(a,d). Most notable, we find many combinations of complexity--entropy values in the lower right part of the plane (high entropy, low complexity), which have been highlighted by rectangles of dotted lines. This region of the plane is challenging for SCMs to distinguish between chaotic time series and random noise~\cite{BorgesAMC2019} and is in our case occupied by values corresponding to chaotic regimes of the logistic map at large $r$ values. The corresponding ambiguity results from the fact that both $\mathcal{H}_O$ and $\mathcal{C}_O$ only take static information in the sense that only pattern frequencies have been considered in their definitions. %The results that  $\mathcal{H}_O$ and $\mathcal{C}_O$ are displaced close to $\mathcal{C}_{max}$ curves have been observed when the embedding $D$ is increased (Fig. \ref{fig:fBmNoise}(a, d)).

For our new pattern transition based SCMs, we find that in the complexity--entropy planes of $\mathcal{C}_W$ versus $\mathcal{H}_W$ and $\mathcal{C}_E$ versus $\mathcal{H}_E$, the results for the logistic map again closely align with the theoretical upper bounds $\mathcal{C}_{max}$. When increasing the embedding dimension, the estimated values approach the theoretical $\mathcal{C}_{max}$ curve more and more closely and essentially cover the left uprising branch of the curve for the SCM based exclusively on transition frequencies (Fig.~\ref{fig:fBmNoise}(b,d)). By contrast, the definition of $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ takes information on both, pattern frequencies and transition frequencies into account, which results in the obtained SCM values aligning even more closely with $\mathcal{C}_{max}$ already at lower embedding dimensions (Fig.~\ref{fig:fBmNoise}(c,f)). The most interesting result is, however, that we do not find any values in the lower right part of the plane, which for the classical SCM based on permutation entropy would present the challenging region of ambiguity. We therefore suggest that it is the deterministic transition behavior among patterns that makes the distinction from random noise.

Based on the results of Fig.~\ref{fig:fBmNoise}, we conclude that the pattern transition behavior encoded in the OPTNs provides novel insights that can be exploited in terms of SCMs that complement their traditional counterparts.

@@

Another conceptual improvement is found for $\mathcal{H}_E$ and $\mathcal{C}_E$ in the parameter range where the logistic map presents period doubling bifurcations, for example, at $r = 3.544$ (Fig.~\ref{fig:bifurcation}(g,h)). It may be noticed that $\mathcal{H}_O$, $\mathcal{C}_O$, $\mathcal{H}_W$ and $\mathcal{C}_W$ all exhibit non-zero values in this parameter range despite a purely periodic dynamics. Furthermore, there are jumps of all four measures at the points of period doubling bifurcations, which have been also reported earlier in Ref.~\cite{BandtPRL2002}. These jumps however are not desirable since the complexity of the dynamics does not change when $r$ passes nay of those points. For the case of the period doubling bifurcation at $r=3.544$ (replacing a period-4 by a period-8 solution), we show in Fig.~\ref{fig:transient} that the observed jump in the entropy and SCM values may be explained by recurrent iterations leading to long transients. \textcolor{red}{I do not fully get this reasoning...} For estimations based on time series of finite length as used in this work, $\mathcal{H}_E$ and $\mathcal{C}_E$ are however not affected in the same way. Certainly, we should not over-interpret their capabilities since numerical inaccuracy would be accumulated in a longer iterative process such that different ordinal patterns are identified. In addition, there are many periodic windows of different periods along the axis of control parameter $r$, which prevents us from using a unique predefined number of iterations as initial transients for time series of different period length. Notably, similar jumps have also been observed when the system bifurcates from a period-3 to a period-6 solution. \textcolor{red}{Yong: give corresponding $r$ value here.}

@@

We have suggested two slightly different ways to consider the effects of pattern transitions in statistical complexity measures, which are based on (i) a globally normalized transition matrix and (ii) a node-wise normalized transition matrix. Here, the complexity measure based on the node-wise normalized matrix considers both, individual pattern frequencies and the frequencies of transitions among patterns. We have shown that the associated two new statistical complexity measures characterize the complexity properties of deterministic chaos successfully and clearly highlight deterministic-chaotic dynamics in the corresponding complexity--entropy planes (Fig.~\ref{fig:fBmNoise}). This finding helps to obtain an improved complexity analysis for time series of both deterministic chaos and stochastic systems, fostering their better understanding. However, further more systematic numerical performance tests employing different types of dynamical systems are required to provide compelling evidence for distinctions between different dynamics beyond the cases presented in the current work, including various models of chaotic maps, long-range correlated stochastic processes like fractional Brownian motion and fractional Gaussian noise, etc.~\cite{BorgesAMC2019,RossoPRE2007}. While this opens several opportunities for follow-up research, we have demonstrated here the practical usefulness of our improved SCMs by employing this framework to time series for experimental fluid flows and ECG recordings, showing distinctive characteristics for different dynamics.

>>>>

As a first step, we consider four representative regimes that can be observed at particular values of the control parameter $r$, corresponding to either periodic dynamics or chaotic dynamics close to band merging, in some laminar state, and at the outer crisis. In all examples, we will consistently use the embedding parameters $D = \tau = 6$. In general, we emphasize that a higher value of $D$ leads to a larger variety of ordinal patterns and, hence, a more robust statistics. On the other hand, for evaluating the statistical properties of either pattern frequencies or transition frequencies, this comes on the cost of requiring increasingly longer time series. In the following, we will use realizations of the logistic map comprising $L = 1 \times 10^6$ iterations.

@@

Due to our choice of $\tau$ and $D$, all SCMs take values of zero for the period-3 series because the temporal distance between each pair of components of the embedding vector covers a multiple of two full periods (cf.~Fig.~\ref{fig:embed}). Other choices of $D$ and $\tau$ would commonly yields non-zero values for the period-3 regime. For the three other cases of band merging, laminar state and outer crisis, the values of all SCMs significantly differ from zero. The pattern frequency based SCM $\mathcal{C}_O$ shows smaller values than the other two measures for all regimes, which implies a positioning in the lower right parts of the complexity--entropy planes in Fig.~\ref{fig:CElogistic}(a). However, for the transition matrix based SCMs, both the estimated entropies $\mathcal{H}_W$, $\mathcal{H}_E$ and associated complexity measures $\mathcal{C}_W$, $\mathcal{C}_E$ show results that are more consistent with the expectations when considering the respective level of chaoticity of the system in those three regimes (i.e., a rising Lyapunov exponent with increasing $r$ in the respective chaotic regimes). In particular, the estimated SCMs take their largest values for $r = 4$ in the outer crisis regime, and the SCMs are the smallest when $r = 3.679$ in the band merging case, while the laminar regime displays intermediate SCM values.

@@

In order to put our previous results into a broader context, we next show the complexity--entropy planes obtained when varying the control parameter $r$ in the logistic map. It is notable that the logistic map has already been widely discussed in the framework of SCMs \cite{RossoPRE2007,MartinPLA2003}, but only for pattern frequency based complexity measures. For each value of $r$, we construct an OPTN from a time series of length $L = 1 \times 10 ^ 6$, after removing the first $1000$ values of transients from the data series. We have repeated the following calculations for embedding parameters $D = \tau \in [2, 7]$.

When only one unique ordinal pattern is identified in a given time series, all SCMs take a value of $0$, which is reasonable since only self-loops are observed. This however happens only if $D$ is an integer multiple of the period for the periodic windows of the logistic map (e.g., Fig.~\ref{fig:embed}(g)). Another practical concern is the data requirements for computing entropies and complexity measures. For a reliable estimation of SCMs, a pragmatic condition on the length $L$ of the time series $L$ as being sufficiently larger than $D!$ \cite{rossoPRL2007,kowalskiPhyD2007} has been satisfied in our setting for $D \leq 7$. In the particular example of logistic map, one may easily generate realizations of longer lengths, which however do not improve the obtained estimates significantly (as we have tested for $L \leq 5 \times 10^6$, not shown).

@@

Figure~\ref{fig:CElogistic} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar when $D$ and $\tau$ are varied in the interval $[3, 7]$. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. By contrast, $\mathcal{C}_{min}$ shows less dependence on $D$.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{(Color online)Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a,d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b,e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c,f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The dashed lines correspond to the maximum $\mathcal{C}_{max}$ and minimum complexity $\mathcal{C}_{min}$ values for a fixed value of the entropy. Panels (a-c) show the results for embedding dimensions $D = 4,5$ and (d-f) for $D = 6$. In panels (a,d), the bottom-right rectangles of dotted red lines highlight the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. }  \label{fig:CElogistic}}
\end{figure*}

First of all, we recover the known result that pattern frequency based SCM values lie around the maximum curve $\mathcal{C}_{max}$, while the displacements of the estimated values from their theoretical upper bounds depend on the embedding dimension $D$ as shown in Fig.~\ref{fig:CElogistic}(a,d). Most notable, we find many combinations of complexity--entropy values in the lower right part of the plane (high entropy, low complexity), which have been highlighted by rectangles of dotted lines. This region of the plane is challenging for SCMs to distinguish between chaotic time series and random noise~\cite{BorgesAMC2019} and is in our case occupied by values corresponding to chaotic regimes of the logistic map at large $r$ values. The corresponding ambiguity results from the fact that both $\mathcal{H}_O$ and $\mathcal{C}_O$ only take static information in the sense that only pattern frequencies have been considered in their definitions.

For our new pattern transition based SCMs, we find that in the complexity--entropy planes of $\mathcal{C}_W$ versus $\mathcal{H}_W$ and $\mathcal{C}_E$ versus $\mathcal{H}_E$, the results for the logistic map again closely align with the theoretical upper bounds $\mathcal{C}_{max}$. When increasing the embedding dimension, the estimated values approach the theoretical $\mathcal{C}_{max}$ curve more and more closely and essentially cover the left uprising branch of the curve for the SCM based exclusively on transition frequencies (Fig.~\ref{fig:CElogistic}(b,d)). By contrast, the definition of $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ takes information on both, pattern frequencies and transition frequencies into account, which results in the obtained SCM values aligning even more closely with $\mathcal{C}_{max}$ already at lower embedding dimensions (Fig.~\ref{fig:CElogistic}(c,f)). The most interesting result is, however, that we do not find any values in the lower right part of the plane, which for the classical SCM based on permutation entropy would present the challenging region of ambiguity. We therefore suggest that it is the deterministic transition behavior among patterns that makes the distinction from random noise.

Based on the results of Fig.~\ref{fig:CElogistic}, we conclude that the pattern transition behavior encoded in the OPTNs provides novel insights that can be exploited in terms of SCMs that complement their traditional counterparts.

%% update but not properly integrated 
We compare the results with fractional Brownian motion (fBm), which has long-range temporal correlations \cite{Mandelbrot1968}. Specifically, for an fBm process the long range of the process is characterized by the Hurst exponent $\alpha$ when positively correlated (persistence) for $1/2 < \alpha < 1$, while suppressed (anti-persistence) for $0< \alpha < 1/2$. $\alpha=1/2$ corresponds to the classical Brownian motion. In an analogy to the previous case of Logistic map of varying control parameter $r$, we generate time series for varying $\alpha \in (0, 1)$ with step size 0.05 and the results are shown in Fig. \ref{fig:CEfBm}.

By the same time, we transform the fBm series into a stationary one by a first-order difference filter, i.e., by considering its increments $x_{i+1}- x_i$. The transformed series is commonly referred to as fractional Gaussian noise (fGn). Notably, fGn retains the long-range correlations and Gaussian probability density function (PDF) from the underlying
fBm process. Note that the time delay embedding issues of both fBm and fGn have been discussed in detail in \cite{Zou2015}. The complexity--entropy planes of fGn are shown in Fig. \ref{fig:CEfGn}.

First, the traditional pattern frequency based SCMs values are reproduced as shown in the Figures \ref{fig:CEfBm}(a, d) and \ref{fig:CEfGn}(a, d) fGn, which agree with that have been reported in \cite{RossoPRE2007,rossoPRL2007}.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fBm.pdf}
\caption{\small{(Color online) Similar with Fig. \ref{fig:CElogistic} but for fractional Brownian motion of Hurst exponent $\alpha \in (0, 1)$ with step size 0.05. (a,d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b,e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c,f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The dashed lines correspond to the maximum $\mathcal{C}_{max}$ and minimum complexity $\mathcal{C}_{min}$ values for a fixed value of the entropy. Panels (a-c) show the results for embedding dimensions $D = 4,5$ and (d-f) for $D = 6$. }  \label{fig:CEfBm}}
\end{figure*}

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fGn.pdf}
\caption{\small{(Color online) Similar with Fig. \ref{fig:CElogistic} but for fractional Gaussian noise Hurst exponent $\alpha \in (0, 1)$ with step size 0.05. (a,d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b,e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c,f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The dashed lines correspond to the maximum $\mathcal{C}_{max}$ and minimum complexity $\mathcal{C}_{min}$ values for a fixed value of the entropy. Panels (a-c) show the results for embedding dimensions $D = 4,5$ and (d-f) for $D = 6$.  }  \label{fig:CEfGn}}
\end{figure*}

@@

Another conceptual improvement is found for $\mathcal{H}_E$ and $\mathcal{C}_E$ in the parameter range where the logistic map presents period doubling bifurcations, for example, at $r = 3.544$ (Fig.~\ref{fig:bifurcation}(g,h)). It may be noticed that $\mathcal{H}_O$, $\mathcal{C}_O$, $\mathcal{H}_W$ and $\mathcal{C}_W$ all exhibit non-zero values in this parameter range despite a purely periodic dynamics. Furthermore, there are jumps of all four measures at the points of period doubling bifurcations, which have been also reported earlier in Ref.~\cite{BandtPRL2002}. These jumps however are not desirable since the complexity of the dynamics does not change when $r$ passes any of those points. For the case of the period doubling bifurcation at $r=3.544$ (replacing a period-4 by a period-8 solution), we show in Fig.~\ref{fig:transient} that the observed jump in the entropy and SCM values may be explained by accumulated numerical errors during the iterations attracted to the periodic-4 points, which yields long transients. \textcolor{red}{I do not fully get this reasoning...} For estimations based on time series of finite length as used in this work, $\mathcal{H}_E$ and $\mathcal{C}_E$ are however not affected in the same way. Certainly, we should not over-interpret their capabilities since numerical inaccuracy would be accumulated in a longer iterative process such that different ordinal patterns are identified. In addition, there are many periodic windows of different periods along the axis of control parameter $r$, which prevents us from using a unique predefined number of iterations as initial transients for time series of different period length. Notably, similar jumps have also been observed when the system bifurcates from a period-3 to a period-6 solution at $r = 3.842$.

@@

We have suggested two slightly different ways to consider the effects of pattern transitions in statistical complexity measures, which are based on (i) a globally normalized transition matrix and (ii) a node-wise normalized transition matrix. Here, the complexity measure based on the node-wise normalized matrix considers both, individual pattern frequencies and the frequencies of transitions among patterns. We have shown that the associated two new statistical complexity measures characterize the complexity properties of deterministic chaos successfully and clearly highlight deterministic-chaotic dynamics in the corresponding complexity--entropy planes (Fig.~\ref{fig:CElogistic}). This finding helps to obtain an improved complexity analysis for time series of both deterministic chaos and stochastic systems, fostering their better understanding. However, further more systematic numerical performance tests employing different types of dynamical systems are required to provide compelling evidence for distinctions between different dynamics beyond the cases presented in the current work, including various models of chaotic maps, long-range correlated stochastic processes like fractional Brownian motion and fractional Gaussian noise, etc.~\cite{BorgesAMC2019,RossoPRE2007}. While this opens several opportunities for follow-up research, we have demonstrated here the practical usefulness of our improved SCMs by employing this framework to time series for experimental fluid flows and ECG recordings, showing distinctive characteristics for different dynamics.
